### 2016-11-13 Manuscript/SW_early_life_telomeres.Rmd
Home eds

#Load and clean data
source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

@@

#####Molecular methods
For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.  Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded. All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Bebbington *et al.* [-@Bebbington2016].

For a subset of first-year birds we had longitudinal data, with an additional sample taken as an adult. For these individuals we calculated the within-individual change in RTL by subtracting early-life RTL from adult RTL (hereafter $\Delta$RTL). Negative values of $\Delta$RTL reflect telomere shortening, while positive values reflect telomere lengthening.

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We tested the relationship between telomere length age and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models with individual ID as a random effect. For the analysis of $\Delta$age we we only included birds who had been sampled at least twice. We then, using the juvenile samples, explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect. We tested how temporal variation in telomere length varied with temporal variation in population density (measured as the number of adult birds on the island at the end of each breeding season) and food availability, using linear mixed models with cohort ID as a random effect.

Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error. We used all birds (adults and juveniles) with at least two telomere measurements (N = `r nrow(subset(ddL,Group == 'Among samples'))` samples from `r length(unique(subset(ddL,Group == 'Among samples')$BirdID))` birds). We then calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample). We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e. completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples). We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age. Secondly, if telomere lengthening is an artifact of measurement error, it is likely to occur more frequently when there is a short follow up time between samples [@Steenstrup2013]. We therefore tested whether telomere lengthening was more likely to occur at short sampling intervals by relating the sample follow up time to both the probability of telomere lengthening (GLM with a binomial error structure, telomere lengthening yes/no as response), and to the dgree of lengthening (LM using only samples where RTL increased over time, $\Delta$RTL as response).

We used linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012]. We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory. We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012]. As random effects we included cohort, territory ID and qPCR plate ID. Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates). For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions in order to obtain more accurate model averaged parameter estimates and confidence intervals.

We used generalised linear mixed models to test whether telomere length/loss were related to survival to adulthood. Models were set up with a binomial error structure, and included RTL or $\Delta$RTL and juvenile age as explanatory variables, along with the same random effects as for the model averaging (see above). We then used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether long-term survival was related to individual-level telomere dynamics. For the cross-sectional data we used lifespan (in years) as survival time, while for the longitudinal data we used remaining lifespan from the second sampling event. We ran these models with cohort as a frailty term to exclude cohort effects, and only included cohorts in which at least 70 percent of individuals had died (all years up to and including 2008).

@@

#####Telomere dynamics and age
Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM with individual ID as random effect: estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,1])` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,2])`; t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`). We found that modelling a non-linear effect age explained almost double the variation in RTL (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than fitting a linear relationship `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`, suggesting that telomere length decreased at the greatest rate in early life. Accordingly, RTL decreased with age at a greater rate in juveniles (estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,2])`) compared to adults (estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,2])`). Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1).

Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (), and was a significantly better model than one including catch year as a random effect. This is indicative of cohort effects on telomere length across the Seychelles warbler lifespan. Furthermore, we found no evidence that the relationship between telomere length and age varied among cohorts (). Rather, cohort-level differences in RTL were consistent over the lifespan of Seychelles warblers.

Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig. 1B). As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest in early life. Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening. To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 1C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).

>>>>

#Load and clean data
source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')
source('Scripts/SW juvenile telomeres IPC models.r')

@@

#####Molecular methods
For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.  Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded. All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Bebbington *et al.* [-@Bebbington2016]. We assessed repeatability of RTL using the rptR package in R [@RDevelopmentCoreTeam2011].

For a subset of birds we had longitudinal data, with an additional samples taken at different ages. For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at timepoint *t* + 1 (hereafter $\Delta$RTL). Negative values of $\Delta$RTL reflect telomere shortening, while positive values reflect telomere lengthening.

#####Statistical analyses
We performed all statistical analyses using R. We first tested the relationship between RTL and age using linear mixed models (LMMs) with individual ID as a random effect, carried out in the lme4 package [@Bates2014]. To test whether the decrease in telomere length with age is greatest in early life, we compared LMMs with age (in months) and log-transformed age, using a likelihood ratio test. We also tested whether telomores shorten with age in individuals with samples from multiple timepoints, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009]. To test how telomere length, and its relationship with age, varied among cohorts, we compared the above LMMs with models that included birth year as an additional random effect term. We also compared a model in which only the intercept (i.e. RTL) was allowed to vary among among cohorts, to one in which the slope of the relationship between RTL and age was also allowed to vary.

Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error. Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample). We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e. completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples). We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age.

We used LMMs to explore how variation in early-life environmental and social conditions influenced telomere length within cohorts. We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, population density (an annual measure estimated from the summer breeding census) and the number of helping and non-helping subordinate birds present in the natal territory. As random effects we included birth year and individual identity. We report model estimates and confidence intervals for all effects included in the full model. As our analysis is exploratory, and because multicollinearity between variables in our full model may have obscured real effects, we also performed model averaging, using the MuMIn package in R [@Barton2012]. Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the plate ID random effect (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates).

@@

#####Telomere dynamics and age
Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`). RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - somehting that was evident from the raw data (Fig. 1). All further models therefore used log-transformed age as a covariate. Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1).

Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (P = `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`). Allowing the slope of the age effect to vary among birth years resulted in a slightly significant improvement on one in which only the intercept was allowed to vary (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`). However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`).

Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig. 1B). As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest in early life. Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening. To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 1C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).

### 2016-11-13 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home eds

Fig3A <- transform(juv,Cohort = factor(LayYear)) %>%
  subset(Agemonths>(1/12)) %>%
  ggplot(aes(y = RTL,x = Cohort))+
  xlab('Cohort') +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12,angle = 90), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))
  annotate('text',x = 0,y = 1.4,label = 'B')

Fig3B <- transform(dd,Cohort = factor(LayYear)) %>%
  ddply(.(Cohort),
        mutate,
        nc = length(which(Agemonths ==1/12))) %>%
  subset(nc > 11) %>%
  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
  xlab('Age (years)') +
  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))
  annotate('text',x = 0,y = 1.4,label = 'B')

>>>>

Fig3A <- transform(juv,Cohort = factor(CatchYear)) %>%
  subset(Agemonths>(1/12)) %>%
  ggplot(aes(y = RTL,x = Cohort))+
  xlab('Cohort') +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12,angle = 90), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))
  annotate('text',x = 0,y = 1.4,label = 'B')

Fig3B <- transform(dd,Cohort = factor(CatchYear)) %>%
  ddply(.(Cohort),
        mutate,
        nc = length(which(Agemonths ==1/12))) %>%
  subset(nc > 11) %>%
  ggplot(aes(y = RTL,x = Agemonths, col = Cohort))+
  xlab('Age (years)') +
  geom_smooth(method = 'lm', formula = y~log(x),se = F) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))
  annotate('text',x = 0,y = 1.4,label = 'B')
